---
title: "Modelling and forecasting deforestation"
author: "Ghislain Vieilledent"
date: "22 octobre 2018"
output:
  html_notebook:
    highlight: tango
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
```

## Computing environment

### Installing Docker

Docker is used to run software packages called "containers". Containers are isolated from each other and bundle their own tools, libraries and configuration files. Containers are created from "images" that specify their precise contents.

To install Docker for your operating system, go to <https://docs.docker.com/install/>. For Windows user, go directly to <https://docs.docker.com/docker-for-windows/install/>.

### Docker image

For our tutorial, we have built a specific Docker image including R and Python:

- R computing environment: R 3.5.1, RStudio, and various R packages (ex. knitr, ggplot2, reticulate).
- Python computing environment: Python 3, and various Python packages (ex. numpy, pandas, matplotlib, and forestatrisk).

Some computations, such as computation by blocks on large raster files, were more easily coded and faster with Python. That's why we decided to use Python as a complementary language to R. The `reticulate` R package allow us to use Python functions from R and thus to work in a single computing environment within RStudio.

### Load the `forestatrisk` Docker image

Download the image from internet with the following command in a terminal:  
`docker pull ghislainv/docker-forestatrisk .` (do not forget the final ".")

Note: If the internet connection is slow, use a USB key and copy the image somewhere on your computer and load it with the following command:  
`docker load -i <path to image tar file>`

Run the container with `docker run -d -e PASSWORD=PWD -p 8787:8787 -v ORIGIN_FOLDER:/home/rstudio ghislainv/docker-forestatrisk`.  

Note:

- Set your own password `PWD`, it cannot be `rstudio`.
- Replace `ORIGIN_FOLDER` with a folder on your local machine where you want to save the R scripts. For example `/c/Users/YOUR_NAME` or `/home/YOUR_NAME`.

Point your browser to `localhost:8787` to access RStudio interface. Log in with `rstudio/PWD`.

### Get the tutorial

In RStudio, open a terminal and clone the GitHub repository locally:  
`git clone https://github.com/ghislainv/forecasting-deforestation-Mada`

Navigate with RStudio in the folder named `forecasting-deforestation-Mada` and open the file named `training.Rmd`. Rmarkdown files (`*.Rmd`) can be used in R to embbed text, code, table and figues inside a unique document. Code are organized into 'chunks' that can be executed step by step. I used this original file to write the tutorial available at <https://ecology.ghislainv.fr/forecasting-deforestation-Mada>.

In the preambule, change the author and date.

### Loading libraries in R

```{r R_libraries}
# Environmental variables
Sys.unsetenv("DISPLAY") # Remove DISPLAY for Python plot

# Libraries
require(reticulate)
require(glue)
require(curl)
require(dplyr)
require(broom)
require(ggplot2)
require(rasterVis)
require(rgdal)

# Source R plotting functions
source("R/dfp_plot.R")
```

### Importing Python modules in R

Specify the Python version to use.

```{r}
use_python("/usr/bin/python3.5")
```

Import the Python modules into R.

```{r importpy}
far <- import("forestatrisk")
patsy <- import("patsy")
sm <- import("statsmodels.api")
smf <- import("statsmodels.formula.api")

```

## Preparing the data-set

### Downloading the geospatial data

```{r download_data}
# Make data directory
if (!dir.exists("data")) {
  dir.create("data")
  dir.create("data/model")
  dir.create("data/mada")
  dir.create("data/validation")
}

# Files
files <- c("fordefor2010.tif", "dist_edge.tif", "dist_defor.tif",
           "altitude.tif", "slope.tif", "aspect.tif",
           "dist_road.tif", "dist_town.tif", "dist_river.tif",
           "sapm.tif", "roads.kml", "towns.kml", "rivers.kml", "sapm.kml")

# Download model data
for (i in files) {
  if (!file.exists(glue("data/model/{i}"))) {
    f <- glue("https://zenodo.org/record/259582/files/{i}")
    curl_download(f, glue("data/model/{i}"), quiet=FALSE)
  }
}

# Download validation data
if (!file.exists("data/validation/for2017.tif")) {
  f <- "http://bioscenemada.cirad.fr/FileTransfer/for2017.tif"
  curl_download(f, "data/validation/for2017.tif", quiet=FALSE)
}

# Download mada outline
if (!file.exists("data/mada/mada38s.shp")) {
  f <- "http://bioscenemada.cirad.fr/FileTransfer/mada38s.zip"
  curl_download(f, "data/mada/mada38s.zip", quiet=FALSE)
  unzip("data/mada/mada38s.zip", exdir="data/mada")
}

```

### Sampling points

```{r sampling}
# Make output directory
if (!dir.exists("output")) {
  dir.create("output")
}

# Training data-set
if (!file.exists("output/sample.txt")) {
  samp <- far$sample(nsamp=10000L, Seed=1234L, csize=10L,
                     var_dir="data/model",
                     input_forest_raster="fordefor2010.tif",
                     output_file="output/sample.txt",
                     blk_rows=1L)
}
samp <- read.table("output/sample.txt", header=TRUE, sep=",")
set.seed(1234)
train <- sample(1:20000, size=10000, replace=FALSE)
data_train <- samp[train,] %>% dplyr::filter(complete.cases(.))
data_valid <- samp[-train,] %>% dplyr::filter(complete.cases(.))
head(data_train)

```

## Deforestation model

### Spatial cells

```{r spatial_cells}
# Spatial cells for spatial-autocorrelation
neighborhood <- far$cellneigh_ctry(raster="data/model/fordefor2010.tif",
                                   vector="data/mada/mada38s.shp",
                                   csize=10L, rank=1L)
nneigh <- neighborhood[[1]]
adj <- neighborhood[[2]]
cell_in <- neighborhood[[3]]
ncell <- neighborhood[[4]]

# Udpate cell number in dataset
cell_rank <- vector()
for (i in 1:nrow(data_train)) {
  cell_rank[i] <- which(cell_in==data_train$cell[i])-1 # ! cells start at zero
}
data_train$cell <- cell_rank

```

### Model formula

```{r formula}
# Formula
data_train$trials <- 1  # Set number of trials to one
formula <- paste0("I(1-fordefor2010) + trials ~ C(sapm) + scale(altitude) + scale(slope) +",
                  "scale(dist_defor) + np.power(scale(dist_defor),2) + ",
                  "scale(dist_edge) + ",
                  "scale(dist_road) + scale(dist_town) + cell")


```

### Fitting model parameters

```{r model}
# Model
mod_binomial_iCAR <- far$model_binomial_iCAR(
  # Observations
  suitability_formula=formula, data=data_train,
  # Spatial structure
  n_neighbors=np_array(nneigh,dtype="int32"), neighbors=np_array(adj,dtype="int32"),
  # Environment
  eval_env=-1L,
  # Chains
  burnin=1000L, mcmc=1000L, thin=1L,
  # Starting values
  beta_start=-99)

```

### Model summary

```{r}
sink(file="output/summary_mod_binomial_iCAR.txt")
print(mod_binomial_iCAR)
sink()
print(mod_binomial_iCAR)
```

### Plot traces and posteriors

```{r plot_with_py, echo=FALSE}
traces_fig <- mod_binomial_iCAR$plot(output_file="output/mcmc.pdf",
                                     plots_per_page=3L,
                                     figsize=c(9,6),
                                     dpi=100)
```

```{r plot_with_r}
require(coda)
mcmc <- as.mcmc(mod_binomial_iCAR$mcmc)
plot(mcmc[,c(1:3,10,11)])
```

## Predicting spatial probability of deforestation

### Resampling the spatial random effects

```{r rho, fig.height=7, message=FALSE, warning=FALSE}
# Get the spatial random effects
rho <- rep(-9999,ncell)  # -9999 will be considered as nodata
rho[cell_in+1] <- mod_binomial_iCAR$rho

# Resample them
far$resample_rho(rho=r_to_py(rho), input_raster="data/model/fordefor2010.tif",
                 output_file="output/rho.tif",
                 csize_orig=10L, csize_new=1L)

# Plot random effects
far$plot$rho("output/rho_orig.tif",output_file="output/rho_orig.png")
far$plot$rho("output/rho.tif",output_file="output/rho.png")

# Plot with R
mada <- rgdal::readOGR(dsn="data/mada",layer="mada38s")
r.rho_orig <- raster("output/rho_orig.tif")
r.rho <- raster("output/rho.tif")
rho_plot(r.rho_orig, mada, output_file="output/rho_orig_ggplot.png",
         quantiles_legend=c(0.025,0.975),width=4.5, height=8)
rho_plot(r.rho, mada, output_file="output/rho_ggplot.png",
         quantiles_legend=c(0.025,0.975),width=4.5, height=8)

```

